---
title: "Lab 8: Mini-Project"
author: "Cecilia Wang (A18625854)"
format: pdf
---

## Background 

In today's class we will apply the methods and techniques clustering and PCA to help make sense of a real world breast cancer FNA biopsy data set. 

## Data import

We start importing our data. It is a CSV file so we will use `read.csv` function.

```{r}
fna.data <- read.csv("WisconsinCancer.csv")
wisc.df <- data.frame(fna.data, row.names=1)
```

```{r}
head(wisc.df,2)
```

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
```

Remove the "diagnosis" line

```{r}
diagnosis <-wisc.df[,1]
```

## Exploratory data analysis

>Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```

>Q2. How many of the observations have a malignant diagnosis?

```{r}
length(grep("M",diagnosis))
# sum(wisc.df$diagnosis==M)
# table(wisc.df$diagnosis)
```

>Q3. How many variables/features in the data are suffixed with _mean?

```{r}
sum(grepl("_mean$", names(wisc.data)))
#length(grep("_mean"),colnames(wisc.data))

```

## Principal Component Analysis

```{r}
# Check column means and standard deviations
colMeans(wisc.data)
apply(wisc.data,2,sd)
```

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data,scale= T)
summary(wisc.pr)
```


> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

44.27%

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

PC3 a captured 72.6% variance

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

PC7 already captured 91% variance

## Interpreting PCA results

```{r}
biplot(wisc.pr)
```

This plot is very hard to read and understand 

Scatter plot: PC1 vs. PC2
```{r}
# Scatter plot observations by components 1 and 2
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

Scatter plot: PC1 vs. PC3

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```


## Variance explained

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <-pr.var/30

# Plot variance explained for each principal component
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```


```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

## Communicating PCA results

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean",1]
```


## Hierarchical clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist)
```

```{r}
plot(wisc.hclust)
abline(h=19.5, col="red", lty=2)
```


```{r}
wisc.hclust.clusters <- cutree(wisc.hclust,k=4)
table(wisc.hclust.clusters, diagnosis)
```

## Combine methods

Here we will take our PCA results and use those as input for clustering. In other words our `wisc.pr$x` scores that we plotted above (the main output from PCA - how the data lie on our new principal component axis/variables) and use a subset of these PCs that captre the most variance as input for `hclust()`.

```{r}
wisc.pr.hclust <-hclust(dist(wisc.pr$x[,1:3]), method = "ward.D2")
plot(wisc.pr.hclust)
```

Cut the dendrogram/tree into two main groups/clusters:

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

I want to know how the clustering in `grps` with values of 1 or 2 correspond the expert `diagnosis`

```{r}
table(grps, diagnosis)
```

My clustering **group 1** are mostly "M" diagnosis (179) and my clustering **group 2** are mostly "B" diagnosis (333)

24 FP
179 TP
333 TN
33 FN

Specificity: TP/(TP+FN)

```{r}
179/(179+33)
```

Specificity: TN/(TN+FP)

```{r}
333/(333+24)
```


```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)
```

## Prediction 

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

>Q. Which of these new patients should we prioritize for follow up based on your results?

Patient 1 should be prioritize. 

