---
title: "Class 5: Data Viz with ggplot"
author: "Cecilia Wang (PID:18625854)"
format: pdf
toc: true
---

## Background

There are a lot of ways to make figures in R. These include so-called "base R" graphics. (e.g. `plot()`) and tones of add-on packages like **ggplot2**. 

For example here we make a same plot with both: 

```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install the package with the command `install.packages()`.

> **N.B.** We never run an install cmd in a quarto code chunk or we will end up re-installing packages many many times- whcih is not we want!

Every time we want to use one of these "add-on" packages we need to load it up in R with`library()` function:

```{r}
library(ggplot2)
ggplot(cars)
```
Every ggplot need at least 3 things:

- The **data**, the stuff you want to plotted
- The **aes**thtics, how the data map to the plot
- The **geom**etry, they type of plot 

```{r}
p <- ggplot(cars)+
  aes(x=speed, y=dist)+
  geom_point()+
  geom_smooth(method = "lm", se = FALSE, color="purple")+
  labs(title = "Stopping distance of old cars", subtitle = "data from the `cars` object", x="Speed (mph)", y="Distance (ft)")
```

render it out
```{r}
p
```

```{r}
p+theme_bw()

```


## Gene expression plot

We can read the input data data the class website

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)

nrow(genes)
```

A first version plot

```{r}
ggplot(genes)+
  aes(Condition1, Condition2) +
  geom_point()
```

```{r}
table(genes$State)
```

Version 2 plot, let's color by `State` so we can see the "u"p and "down" significant genes compared to all the unchanging genes. 

```{r}
ggplot(genes)+
  aes(Condition1, Condition2, col=State) +
  geom_point()
```

Version 3 plot, let's modify the colors to something we like:

```{r}
ggplot(genes)+
  aes(Condition1, Condition2, col=State) +
  geom_point() +
  scale_color_manual(values = c("Blue","grey","red"))+
  labs(x="Control (no drug)", y="Drug",
       title = "Gene Expression Change upon GLP-1 Drug")+
  theme_linedraw()
```

## Going Further

Let's have a look at the famous **gapminder**

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"
gapminder <- read.delim(url)
```

```{r}
head(gapminder,5)
```

Version 1 Plot

```{r}
ggplot(gapminder)+
  aes(x=gdpPercap, y=lifeExp)+
  geom_point(alpha=0.4)
```

Version 2 plot:

```{r}
ggplot(gapminder)+
  aes(x=gdpPercap, y=lifeExp,col=continent)+
  geom_point(alpha=0.3)
```

Let's "facet" by continent rather than big hot mess above

```{r}
ggplot(gapminder)+
  aes(x=gdpPercap, y=lifeExp,col=continent)+
  geom_point(alpha=0.3)+
  facet_wrap(~continent)
```

## Custom Plots

How big is this gapminder dataset?

```{r}
nrow(gapminder)
```

I want to "filter" down to a subset of this data. I will use **dplyr** package to help me. 

 First I need to install and load it up...
 
```{r}
library(dplyr)

gapminder_1977 <- gapminder %>% filter(year==1977)
head(gapminder_1977)
gapminder_2007 <- gapminder %>% filter(year==2007)
head(gapminder_2007)
```

```{r}
filter(gapminder_2007, country=="United States")
```

> Q. Make a plot compring 1977 and 2007 for all countries. 

```{r}
gap_compare <- filter(gapminder,year %in% c(1977,2007))

ggplot(gap_compare)+
  aes(x=gdpPercap, y=lifeExp, col=continent)+
  geom_point(alpha=0.9)+
  facet_wrap(~year)
  
```










